---
title: "SampleQC: Generating plots for paper"
author:
- name: Will Macnair
  affiliation: 
  - &IMLS Institute for Molecular Life Sciences, University of Zurich, Switzerland
  - Swiss Institute of Bioinformatics (SIB), University of Zurich, Switzerland
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  workflowr::wflow_html:
    code_folding: show
    toc: true
    toc_float: true
    number_sections: false
---

# Setup / definitions

## Libraries

```{r setup_knitr, include=FALSE}
library('BiocStyle')
knitr::opts_chunk$set( autodep=TRUE, cache=TRUE, dev='png' )
# wflow_build(files='analysis/qc04_real.Rmd', view=F, verbose=T, delete_cache=F)
```

```{r setup_libs, collapse=FALSE, message=FALSE, warning=FALSE, cache=FALSE}
library('SampleQC')
```

## Inputs

```{r setup_input}
# where is snRNAseq file?
qc_f        = 'data/snRNAseq/sn_qc_dt_anon_2021-07-25.txt.gz'
types_f     = 'data/snRNAseq/sn_types_dt_2021-07-25.txt.gz'
qc_obj_f    = 'data/snRNAseq/sampleQC_obj_2021-07-25.rds'

qc_names    = c('log_counts', 'logit_mito', 'splice_ratio')
annots_disc = c('donor_id', 'condition', 'batch', 'source')
annots_cont = c('med_splice')
```

## Helper functions

```{r setup_helpers, cache=FALSE}
source('code/qc03_sims.R')
source('code/qc04_real.R')
```

## Outputs

```{r setup_output}
# where to save?
save_dir  = 'output/qc04_real'
if (!dir.exists(save_dir))
  dir.create(save_dir)

# annotations to save
disc_ls         = c('group_id', 'batch', 'mito_cat', 'N_cat')
cont_ls         = c('med_counts', 'med_splice')
annot_ls        = c(disc_ls, cont_ls)

# exclude clusters with above this splice_ratio
splice_cut      = 3

# what figs to save?
formats         = c('png', 'pdf')
outs_scater_f   = file.path(save_dir, 'outs_scater_bio.%s')
outs_distns_f   = file.path(save_dir, 'outs_distns_bio.%s')
kept_distns_f   = file.path(save_dir, 'kept_distns_bio.%s')
maha_qqs_f      = file.path(save_dir, 'mahalanobis_qq_plot_bio.%s')
outlier_qcs_f   = file.path(save_dir, 'outlier_qcs_bio.%s')

seed            = 20200725
set.seed(seed)
```

# Load inputs

# Processing / calculations

```{r load_qc}
# original QC variables
qc_all      = qc_f %>% fread %>%
  .[, mito_prop := plogis(logit_mito) ]
# variables selected for SampleQC, plus annotations
qc_dt       = make_qc_dt(qc_all, qc_names = qc_names) %>%
  .[, med_splice := median(splice_ratio), by=sample_id]
```

```{r load_celltypes}
types_dt    = fread(types_f)
```

```{r run_SampleQC}
# K_list    = c(4, 4, 3, 2, 1, 3, 4, 4, 4)
K_list    = c(3, 4, 1, 4, 2, 1, 3, 4)
# K_list    = c(1, 1, 1, 1, 1, 1, 1, 1)
if (file.exists(qc_obj_f)) {
  qc_obj    = readRDS(qc_obj_f)
} else {
  # find sample groups
  set.seed(seed)
  qc_obj    = calc_pairwise_mmds(qc_dt, qc_names, 
    annots_disc = annots_disc, annots_cont = annots_cont, 
    n_cores = 16)
  print(table(colData(qc_obj)$group_id))

  # fit model
  set.seed(seed)
  qc_obj    = fit_sampleqc(qc_obj, K_list = K_list, mcd_alpha = 0.3)

  # save outputs
  make_sampleqc_report(qc_obj, save_dir, 'snRNAseq')

  # specify outlier cluster
  saveRDS(qc_obj, file = qc_obj_f)
}
```

```{r calc_outliers_dt}
# extract component means
group_ids   = levels(colData(qc_obj)$group_id)
beta_ks     = group_ids %>%
  lapply(function(g) {
    beta_k  = metadata(qc_obj)$fit_list[[g]]$beta_k
    mu_0    = metadata(qc_obj)$fit_list[[g]]$mu_0
    beta_k  = sweep(beta_k, 2, mu_0, '+')
    assert_that( all(abs(colMeans(beta_k) - mu_0) < 1e-10) )
    return(beta_k)
  })

# get clusters with positive splice ratio
splice_idx    = qc_names == 'splice_ratio'
exc_clusters  = lapply(beta_ks, 
  function(beta_k_mat) which(beta_k_mat[, splice_idx] > splice_cut)
  ) %>% setNames(group_ids)
exc_clusters  = exc_clusters[ sapply(exc_clusters, function(x) length(x) > 0) ]

# get outliers
outliers_dt   = get_outliers(qc_obj, exc_clusters = exc_clusters)
```

```{r calc_scater_miqc}
# calc scater outliers
scater_dt = calc_scater_dt(qc_all)
set.seed(123)
miqc_dt   = calc_miqc_dt(qc_all)
```

```{r join_all_outliers}
all_outliers  = make_all_outliers(outliers_dt, scater_dt, miqc_dt)
outlier_qcs   = make_outlier_qcs(outliers_dt, scater_dt, miqc_dt, qc_all)
# join outliers to types
outlier_bias  = all_outliers %>%
  merge(types_dt, by = c('sample_id', 'cell_id')) %>%
  .[, .N, by = .(sample_id, type_broad, method, outlier)] %>%
  .[, prop := N / sum(N), by = .(sample_id, type_broad, method)]
```

# Analysis

## Embeddings of samples{.tabset}

```{r plot_embeddings_disc, fig.height = 9, fig.width = 8, results = 'asis', cache = FALSE}
for (sel_embed in c('UMAP', 'MDS')) {
  cat('### ', sel_embed, '\n')
  g_list  = lapply(seq_along(annot_ls), function(ii) {
    annot_i   = annot_ls[[ii]]
    var_type  = ifelse(annot_i %in% disc_ls, 'discrete', 'continuous')
    SampleQC:::.plot_one_embedding(qc_obj, annot_i, var_type, sel_embed) +
      labs( title=LETTERS[[ii]] )
    })
  g       = wrap_plots(g_list, ncol=2)
  print(g)
  png_f   = sprintf('%s/real_qc_embed_disc_%s.png', save_dir, sel_embed)
  ggsave(png_f, g, h=9, w=8)
  cat('\n\n')
}
```

## Plot `SampleQC` model fits and outliers over QC biaxials

This takes a while to render, so for speed I've set `eval = FALSE` - feel free to remove that call.

```{r plot_histograms_split, fig.height=6, fig.width=7, results='asis', eval = FALSE}
group_list  = levels(qc_obj$group_id)
for (g in group_list) {
  cat('### ', g, '{.tabset}\n')
  # which samples?
  samples_g   = sort(colData(qc_obj)$sample_id[ colData(qc_obj)$group_id == g ])
  # # just do a handful
  # samples_g   = sample(samples_g, min(5, length(samples_g))) %>% sort
  for (s in samples_g) {
    cat('#### ', s, ' \n')
    g_fit   = plot_fit_over_biaxials(qc_obj, s)
    g_out   = plot_outliers(qc_obj, s)
    g     = g_fit / g_out
    print(g)
    cat('\n\n')
  }
}
```

## Plot outliers

```{r plot_outliers, fig.height = 8, fig.width = 10}
# assemble plot
s       = 'sample017'
g_fit   = plot_fit_over_biaxials(qc_obj, s) + 
  labs( title = 'A. SampleQC fits' )
g_maha  = plot_maha_dists(qc_obj, s) + 
  labs( title = 'B. Mahalanobis distances' )
g_qc    = plot_outliers(qc_obj, s, outliers_dt) + 
  labs( title = 'C. SampleQC outliers' )
g_scat  = plot_outliers(qc_obj, s, scater_dt) + 
  labs( title = 'D. scater outliers' )
g_miqc  = plot_outliers(qc_obj, s, miqc_dt) + 
  labs( title = 'E. miQC outliers' )
g       = list(g_fit, g_maha, plot_spacer(), g_qc, g_scat, g_miqc) %>% 
  wrap_plots(ncol = 2, byrow = FALSE)

# save files
for (f in formats)
  ggsave(sprintf(outs_scater_f, f), g, h = 8, w = 10)

# plot
print(g)
```

## Plot Mahalanobis distances

```{r plot_maha_qq, fig.height = 5, fig.width = 6}
set.seed(seed)
sel_samples   = qc_obj$sample_id %>% sample(6) %>% sort
mahas_dt      = calc_mahas_dt(qc_obj)

# do plot
g             = plot_maha_qqs(mahas_dt, sel_samples)
for (f in formats)
  ggsave(sprintf(maha_qqs_f, f), g, h = 5, w = 8)
print(g)
```

## Plot QC metrics for outliers under each method

```{r plot_outlier_metric_distributions, fig.height = 8, fig.width = 7}
# do plot
set.seed(seed)
sel_samples   = qc_obj$sample_id %>% sample(6) %>% sort
for (f in formats)
  ggsave(sprintf(outlier_qcs_f, f), g, h = 8, w = 7)
print(g)
```

## Plot exclusions under alternative methods

```{r plot_outlier_bias, fig.height = 5, fig.width = 6}
g = ggplot(outlier_bias[ outlier == TRUE ]) +
  aes( x = prop, colour = method, fill = method ) +
  # aes( x = type_broad, y = prop, colour = method, fill = method ) +
  # stat_ecdf() +
  geom_density( alpha = 0.5 ) +
  # coord_flip() +
  # scale_y_continuous( breaks = pretty_breaks() ) +
  scale_x_continuous( breaks = pretty_breaks() ) +
  scale_colour_manual( values = method_cols ) +
  scale_fill_manual( values = method_cols ) +
  facet_grid( type_broad ~ ., scales = 'free_y' ) +
  # facet_wrap( ~ type_broad ) +
  theme_bw() + theme( panel.grid = element_blank() )
print(g)
```
